---
title: "Possible features for survey analysis"
author: "Artem Larionov"
date: "September 4, 2016"
output: html_document
abstract: In this article I've applied a few techniques of text analysis to see which ones would be useful to analyze surveys.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
setwd('~/articles/20160904-survey-analysis')

library(dplyr)
library(syuzhet)
library(tm)
library(RWeka)
library(SnowballC)  
library(stringr)
library(wordcloud)
library(RColorBrewer)
library(plyr)
library(ggplot2)
library(easyGgplot2)

prepareString <- function (x) {
  x <- removeWords(x, stopwords())
  x <- iconv(x, 'latin1', 'ASCII', sub='')
  x <- tolower(x)
  x <- removePunctuation(x)
  x <- removeNumbers(x)
  x <- stripWhitespace(x)
  str_trim(x)
}

buildDTMatrix <- function (corpus, min, max, language = 'english', weighting = weightTf) {
  DocumentTermMatrix(
    corpus, 
    control = list(
      tokenize = function(x) {
        NGramTokenizer(x, Weka_control(min = min, max = max))
      },
      language = language,
      stemWords = FALSE,
      wordLengths=c(3,Inf),
      weighting = weighting
    )
  )
}

question.text <- function (id) { questions[questions$id == id,]$body }
topic.id      <- function (id) { questions[questions$id == id,]$topic_id }
topic.title   <- function (id) { topics[topics$id == id,]$title } 

topics    <- read.csv('topics.csv', stringsAsFactors = FALSE)
questions <- read.csv('questions.csv', stringsAsFactors = FALSE) 
answers   <- read.csv('answers.csv', stringsAsFactors = FALSE)

answers <- subset(answers, company_id == 3)
answers$question <- sapply(answers$poll_id, question.text)
answers$topic_id <- as.factor(sapply(answers$poll_id, topic.id))
answers$topic    <- sapply(answers$topic_id, topic.title)
answers$answer_sentiment <- get_nrc_sentiment(answers$text)[1:8]
answers$question_sentiment <- get_nrc_sentiment(answers$question)[1:8]

poll_87 <- subset(answers, poll_id == 87)
poll_32 <- subset(answers, poll_id == 32)
poll_59 <- subset(answers, poll_id == 59)
poll_62 <- subset(answers, poll_id == 62)

custom_questions <- subset(questions, topic_id == 7 & company_id == 3)
```

# Introduction

Right now I'm not going to focus on details of the implementation, but if you are interested, you can find the source code [here](https://github.com/alarionov/articles/blob/gh-pages/20160904-survey-analysis/20160904-survey-analysis.Rmd).

Let's start with word clouds for questions and answers:

```{r}
par(mfrow=c(1,2))
wordcloud(answers$question, random.order = F, colors = brewer.pal(6, 'Dark2'), main = 'Word Cloud for Questions')
wordcloud(answers$text, random.order = F, colors = brewer.pal(6, 'Dark2'), main = 'Word Cloud for Answers')
```

As we could expect, the questions are focused on "what" and "how" about "company" and "work".
Answers look quite positive with a big "good" in the middle, the biggest "know" is probably related to "don't" right above it :). It might be useful to find those answers and questions to understand whether the employees don't know something important, or just haven't gotten familiar with the survey tool yet :). 

# Word frequency

Let's dig deeper. Word frequency can be useful for some kinds of questions because it will summarize all answers into the most common words.

```{r}
corpus <- Corpus(VectorSource(paste(poll_62$text, collapse = ' ')))
corpus <- tm_map(corpus, prepareString)
corpus <- tm_map(corpus, PlainTextDocument)
unigram_sparse_dtm <- buildDTMatrix(corpus, 1, 1)
matrix <- as.matrix(unigram_sparse_dtm)
uni_sorted <- matrix[1,order(matrix[1,])]
par(mar=c(5,5,4,2) + 0.1, mfrow=c(1,1))
barplot(tail(uni_sorted, n =10), horiz = TRUE, cex.names = 0.7, las = 1, main = questions[questions$id == 62,]$body)
```

Looks good, but what is that "disturb" in the middle?

```{r}
poll_62$text[grep('disturb', poll_62$text)]
```

As we can see, this approach doesn't consider the context of the answer (and the question), so it's always good to check what is hiding behind the numbers. 

# Sentiments

So, word frequency gave us some keywords for our answers, let's check which ones of them are positive.

```{r}
sentiments <- get_nrc_sentiment(c('good', 'friendly', 'work', 'just', 'fine'))[9:10]
rownames(sentiments) <- c('good', 'friendly', 'work', 'just', 'fine')
sentiments
```

We can do the same for answers to the question.

```{r}
barplot(sort(colSums(poll_62$answer_sentiment)), horiz = TRUE, cex.names = 0.7, las = 1, main= questions[questions$id == 62,]$body)
```

Or even for the entire topic.

```{r}
sentiments_by_topic <- do.call(rbind.data.frame, by(answers, answers$topic_id, function(x) colSums(x$answer_sentiment), simplify = F))
colnames(sentiments_by_topic) <- colnames(answers$answer_sentiment)
rownames(sentiments_by_topic) <- sapply(rownames(sentiments_by_topic), topic.title)

for (i in c(1,4,5)) {
  barplot(
    as.matrix(sort(sentiments_by_topic[i,])), 
    horiz = TRUE, 
    cex.names = 0.7, 
    las = 1, 
    main = rownames(sentiments_by_topic)[i]
  )
}
```

And for the whole company.

```{r}
barplot(sort(colSums(answers$answer_sentiment)), horiz = TRUE, cex.names = 0.7, las = 1, main="Commulative sentiments for all answers")
```

We can also combine word frequencies and sentiments.

```{r}
corpus <- Corpus(VectorSource(paste(answers$text, collapse = ' ')))
corpus <- tm_map(corpus, prepareString)
corpus <- tm_map(corpus, PlainTextDocument)
unigram_sparse_dtm <- buildDTMatrix(corpus, 1, 1)
matrix <- as.matrix(unigram_sparse_dtm)
uni_sorted <- matrix[1,order(matrix[1,])]
sentiments <- get_nrc_sentiment(names(uni_sorted)) * uni_sorted
sentiments <- sentiments[1:8]

for (i in 1:nrow(sentiments)) {
  row = sentiments[i,]
  if (sum(row) > 0) {
    row = row / sum(row)
    sentiments[i,] = row 
  }
}

indicies = list()
for (name in colnames(sentiments)) {
  indicies[[name]] <- as.numeric(rownames(sentiments[sentiments[name] >= 0.5,]))
}

par(mfrow=c(2,3))
for (name in names(indicies)) {
  if (length(indicies[[name]])) {
    barplot(tail(uni_sorted[indicies[[name]]], n = 5), horiz = T,cex.names = 0.7, las = 1, main = name)
  }
}
```

We can see that answers are mostly positive, but is it getting better or worse?
Let's check how sentiments are changing over time.

```{r}
sentiments_by_survey <- do.call(
  rbind.data.frame, 
  by(
    answers, 
    answers$survey_template_session_id, 
    function(x) colMeans(x$answer_sentiment), 
    simplify = F
  )
)
colnames(sentiments_by_survey) <- colnames(answers$answer_sentiment)
par(mfrow=c(2,4))
for (i in 1:8) {
  plot(
    1:nrow(sentiments_by_survey),
    sentiments_by_survey[[i]], 
    type = 'l',
    cex.names = 0.7,
    xlab = '# of surveys',
    ylab = colnames(sentiments_by_survey)[i],
    main = colnames(sentiments_by_survey)[i]
  )
}
```

We don't have enough data to be sure but it looks like positive sentiments are used more actively.

# Conlusion

These techniques could be used as a summary/overview of а big amount of text, but we also need to think about tools for more detailed analysis.